---
title: "Spatial Visualization of Oil Spill Events in California"
description: |
  These maps, which display locations and numbers of oil spills in California for the year 2008, were created in R-Studio 1.4.
author:
  - name: Kelsie Fronheiser
    url: {}
date: 02-22-2021
output:
  distill::distill_article:
    self_contained: false
    theme: lumen
    code_folding: hide
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
```

The spatial data used in this report was collected from the [Oil Spill Incident Tracking database](https://gis.data.ca.gov/datasets/7464e3d6f4924b50ad06e5a553d71086_0/data) by the California Department of Fish and Game Office of Spill Prevention & Response, published on July 23, 2009.  

```{r}
oil_spills <- read_sf("ds394.shp")
ca_counties <- read_sf(here("_posts", "2021-03-14-visualization-of-oil-spill-events-in-california", "ca_counties","CA_Counties_TIGER2016.shp"))

ca_subset <- ca_counties %>% 
  select(NAME, ALAND) %>% 
  rename(county_name = NAME, land_area = ALAND)

#ca_subset %>% st_crs()
#oil_spills %>% st_crs()

oil_spills <- st_transform(oil_spills, 3857)

#oil_spills %>% st_crs()
```

## Interactive Map of Inland Oil Spill Events in California
```{r}

# Set the viewing mode to "interactive":
tmap_mode(mode = "view")

tm_shape(ca_subset) +
  tm_borders() +
  tm_shape(oil_spills) +
  tm_dots(col = "red1") +
  tmap_style("natural")

```


**Figure 1.** California oil spill events for the year 2008 are displayed by red points. Click on a point to see information about the individual spill event.

```{r}
ca_oil <- ca_subset %>% 
  st_join(oil_spills)

oil_counts <- ca_oil %>% 
  count(county_name)
```

## Choropleth Map of Inland Oil Spill Events by California County
```{r}
ggplot(data = oil_counts) +
  geom_sf(aes(fill = n), color = "black", size = 0.1) +
  scale_fill_gradientn(colors = c("white","orange","red")) +
  theme_void() +
  labs(fill = "Number of oil spills")
```

**Figure 2.** California oil spill events for the year 2008 are summed by county, with the abundance of spills displayed by a gradient from zero spills in white to 500+ spills in red. The county with the highest number of oil spills in 2008 was Los Angeles County.